---
title: "DiscreteCharacters"
author: "Brian O'Meara"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

For this exercise, explore looking at discrete character models. Note: they are not "discreet" character models -- they're actually pretty noisy. To check your work, click on the `knit` button in RStudio. Make sure to change `eval=FALSE` to `eval=TRUE` to run the code when `knit`ting.

These are useful packages in this area (far from exhaustive list -- see the [CRAN task view](https://cran.r-project.org/web/views/Phylogenetics.html) for more). Rather than have to worry about do you have dependencies installed, etc. this will use the yearn package to load a package if you have it, or install and load it if you don't. In real life, you should have all the packages you need already installed and call with a `library()` call; best practices are evolving to use things like [https://rstudio.github.io/packrat/](packrat) to keep consistent package versions throughout an analysis.


```{r setup, include=FALSE, echo=FALSE}
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
install.packages("yearn")
yearn::yearn(ape)
yearn::yearn(geiger)
yearn::yearn(phytools)
yearn::yearn(phangorn)
yearn::yearn(corHMM)
```

You'll need to get data into R in some way. Look at other phylometh assignments for how to get trees and data.


```{r, eval=TRUE}
MyTree <- read.tree("MangroveNewick2.tre")
data1 <- read.csv(file="MangroveDiscrete.csv", stringsAsFactors=FALSE) #death to factors.

#Tree was taken using TreeSnatcher from Reid, D. G., P. Dyal, and S. T. Williams. "Global diversification of mangrove fauna: a molecular phylogeny of Littoraria (Gastropoda: Littorinidae)." Molecular phylogenetics and evolution 55, no. 1 (2010): 185-201.


#My discrete data is Obligate Mangrove-Dwelling = 1 , Non-obligates = 0
```

Data are often not right in some way. They might not match the taxa in your tree, there may be missing data, etc. `geiger::treedata` is a great function for getting a tree and data that match, but your data may need other cleaning. Do it as a function so it's repeatable.

```{r, eval=TRUE}

mangrove.vector1<-data1[,2]
names(mangrove.vector1)<-data1[,1]
name.check(MyTree, mangrove.vector1)

CleanData <- function(MyTree, mangrove.vector1) {
  geiger::treedata(MyTree, mangrove.vector1)
}

# Now write the code to use CleanData() to actually clean your data

```

It's critically important to LOOK at what you have. Are there weird values? Has the match between taxa and state gone correctly? Do you think you have binary data, but there's actually only state 1? Especially as data sets grow (yay), and are assembled using scripts rather than by error-prone, non-reproducable hands (double yay), scientists are increasingly less likely to deeply look at our data. That's bad -- don't be that person.

```{r, eval=TRUE}
VisualizeData <- function(MyTree, mangrove.vector1) {
	#Important here is to LOOK at your data before running it. Any weird values? Does it all make sense? What about your tree? Polytomies?

plot(MyTree)

#lookin' hot

}
```

First, let's use parsimony to look at ancestral states:

```{r, eval=TRUE}

clean.data.2<-as.matrix(mangrove.vector1,ncol=1)
clean.phyDat<-as.phyDat(clean.data.2,type="USER",levels=c(0,1))
anc.p <-phangorn::ancestral.pars(MyTree,clean.phyDat)
plotAnc(MyTree,anc.p,1)



```

Do you see uncertainty? What does it mean?

My tree looks fairly certain, other than the base node.

Now, plot the likelihood estimates.

```{r, eval=TRUE}
anc.ml <- ancestral.pml(pml(MyTree, clean.phyDat), type="ml")
plotAnc(tree, anc.ml, 1)
```

How does this differ from parsimony?

Much more uncertainty in the liklihood estimates. Every node beyond the tips has an equal liklihood of being either 1 or 0 state...interesting

Why does it differ from parsimony?

Parimony is looking for the tree with the fewest changes in character state. Liklihood is using the probability of all character changes.

What does uncertainty mean?

Which node had what character state? We're unsure about this...so this is our uncertainty

Now, to the biological questions. For many of these, `corHMM` will be a useful package. Do the following analyses:

1. How can you estimate transition rates between states? Do it.
```{r, eval=TRUE}

devtools::install_github("thej022214/corHMM",dependencies=FALSE) #Analysis of Binary Character Evolution v1.24

data2<-data.frame(taxon=names(mangrove.vector1),trait=unlist(mangrove.vector1),stringAsFactors=FALSE)
cordat<-corHMM(MyTree,data2,rate.cat=2,node.states="none")

```
2. How could you examine if transition rates are equal?

#Look at cordat


3. Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change?

LewisMK<-lewisMkv(MyTree,data2)

#so I'm pretty sure this is correct yet I'm receiving the error below vvv

#"Error in apply(tmp, 1, which.max) : dim(X) must have a positive length"

4. How could you test order of state evolution?

```{r,eval=TRUE}

x0<-corHMM(MyTree,data2,rate.cat=2,node.states="none",root.p=1)
x1<-corHMM(MyTree,data2,rate.cat=2,node.states="none",root.p=0)
```

